---
title: "Kinetic model comparison for post-export RNA conversion"
author: "Luca Faretra, Francesco Napolitano, Massimo Pancione, Luigi Cerulo"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Kinetic model comparison for post-export RNA conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 72)
```

# Introduction

`postexportKinetics` fits a compartment-resolved kinetic model to time
courses of nuclear and cytoplasmic, unprocessed and processed RNA. It asks
whether an additional post-export conversion component improves the fit to
the data beyond what the model explains without it.

The package implements the method of the accompanying manuscript. Its
numerical core is ported verbatim from the frozen manuscript implementation
(tag `manuscript-revision-v1.0`) and is regression-tested against it.

This vignette works through one small synthetic dataset:

- data formatting and validation;
- fitting, testing, interpretation, multiple testing and exploratory ranking;
- plotting, simulation and design diagnostics;
- limitations.

# Installation

Once the package has been accepted and released through Bioconductor, it
will be installed with `BiocManager`. This route does not work before
acceptance:

```{r install-bioc, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("postexportKinetics")
```

In the current development state, install from GitHub, either the released
version 0.1.0 or the development line:

```{r install-github, eval = FALSE}
# install.packages("remotes")
remotes::install_github("bioinformatics-sannio/postexportKinetics@v0.1.0")
remotes::install_github("bioinformatics-sannio/postexportKinetics",
                        build_vignettes = TRUE)
```

Load the package:

```{r load}
library(postexportKinetics)
```

# The four-state model

Each event (for example a retained-intron event) is described by four
states:

| State | Compartment | Processing |
|---|---|---|
| `N`   | nuclear     | unprocessed |
| `N_s` | nuclear     | processed   |
| `C`   | cytoplasmic | unprocessed |
| `C_s` | cytoplasmic | processed   |

with dynamics

```
dN/dt   = R(t) - (sigma_n + tau) N
dN_s/dt = sigma_n N - tau_s N_s
dC/dt   = tau N - (sigma_c + alpha) C
dC_s/dt = tau_s N_s + sigma_c C - alpha_s C_s
```

The rates are:

- `R(t)`: transcription;
- `sigma_n`: nuclear processing;
- `tau` and `tau_s`: export of unprocessed and processed RNA;
- `alpha` and `alpha_s`: cytoplasmic decay;
- `sigma_c`: the **post-export conversion rate** from `C` to `C_s`.

Two nested models are compared:

- the **null model**, with `sigma_c = 0`: no post-export conversion
  component;
- the **full model**, with `sigma_c >= 0`: a non-negative post-export
  conversion rate.

`sigma_c` is a *phenomenological* parameter. A positive estimate, or a small
p-value, indicates kinetic patterns consistent with an additional
post-export conversion component *within this model*. It does not identify a
unique molecular mechanism, and it is not by itself evidence of cytoplasmic
splicing.

Transcription follows a schedule with a transcriptional onset and a common
intervention time `t_star`. After `t_star`, transcription stops completely
(shutoff) in the analyses shown here.

# Example data and input format

The package ships a small **synthetic** dataset, `postexport_example`. It was
simulated with the package's simulator (see `?postexport_example` and
`data-raw/postexport_example.R` in the source repository):

- eight events: four simulated with `sigma_c > 0` (`alt_*`) and four with
  `sigma_c = 0` (`null_*`);
- complete shutoff at `t_star = 332` min;
- 5 time points and 5 destructive replicates per time point;
- RNA-seq-type noise.

The design was selected from the manuscript benchmark table before any data
were generated. It is the calibrated design (Type-I error interval
containing 0.05) with the highest benchmark power among small shutoff
designs, and that power is only about 0.24. The simulation seed was fixed a
priori; neither the design nor the data were chosen by searching over
seeds. The example therefore shows realistic uncertainty and does not
guarantee detection.

```{r data}
data(postexport_example)
head(postexport_example)
```

The canonical **wide** format has one row per destructive sample, with
columns `event`, `time`, `replicate`, `N`, `N_s`, `C` and `C_s`.

- Replicate labels identify samples *within* a time point. They are never
  used to pair samples across time points.
- `time_unit` is required, because all rates, including `sigma_c`, are per
  this unit.

```{r data-object}
x <- postexport_data(postexport_example, time_unit = "min")
x
```

The **long** format has one row per sample and state, with columns
`compartment` (`nuclear`/`cytoplasmic`), `state`
(`unprocessed`/`processed`) and `abundance`. The same observations are
installed as a CSV file:

```{r data-long}
path <- system.file("extdata", "postexport_example_long.csv",
                    package = "postexportKinetics")
long <- read.csv(path)
head(long, 4)
x_long <- postexport_data(long, time_unit = "min", format = "long")
```

Different column names can be declared with `columns`, for example
`columns = c(event = "event_id", time = "minutes")`.

For retained-intron events, the mapping of inclusion to *unprocessed* and
skipping to *processed* is the user's responsibility. It is not inferred.

# Validation

`postexport_data()` validates its input. `validate_postexport_data()` reports
the findings without constructing an object, and never modifies the data.

```{r validate}
validate_postexport_data(postexport_example, t_star = 332)
```

- **Errors:** missing states, missing or non-finite values, duplicated
  samples, fewer than two time points, or no time point with at least two
  replicates. Missing values are never set to zero or imputed.
- **Warnings:** negative abundances and an intervention time outside the
  sampled window.
- **Notes:** diagnostic information.

# Fitting the full and null models

`fit_postexport_model()` fits both models without bootstrap. The estimation
uses:

- trapezoidal interval balances between consecutive time points;
- within-time covariance estimation with shrinkage and an eigenvalue floor;
- covariance propagation and whitening;
- non-negative least squares.

The intervention time `t_star` is always an explicit argument.

```{r fit}
sel <- postexport_example$event == "alt_3"
one <- postexport_data(postexport_example[sel, ], time_unit = "min")
fit <- fit_postexport_model(one, t_star = 332)
fit
```

`summary()` shows the full and null coefficients, the residual sums of
squares, the boundary information and numerical diagnostics such as
condition numbers and ranks.

```{r fit-summary}
summary(fit)
```

# Inference: bootstrap test of `sigma_c = 0`

`test_postexport_conversion()` compares the models with the statistic
`T = max(0, RSS0 - RSS1)`, a difference of whitened non-negative
least-squares residual sums of squares. This is **not** a likelihood-ratio
test.

The null distribution of `T` comes from a replicate-level generative
bootstrap under the fitted null model. The interval system is rebuilt in
every bootstrap replicate, and the p-value uses the add-one rule
`(1 + #{T* >= T_obs}) / (B + 1)`. The alternative is one-sided,
`sigma_c > 0`.

Several events can be tested in one call. They are processed sequentially,
and each event gets its own named seed:

```{r test}
events <- unique(x$event)
seeds <- setNames(seq_along(events), events)
res <- test_postexport_conversion(
    x, t_star = 332,
    control = postexport_control(B = 499, seed = seeds))
res
```

`B = 499` is used only to keep this vignette fast and is not recommended
for final scientific analysis. The smallest attainable p-value is
`1 / (B + 1)`, here 0.002. The package default is `B = 1999`, and the
manuscript's final real-data analyses used larger values.

With an explicit seed, the caller's random-number state is restored
afterwards. With `seed = NULL`, the global stream is used and consumed.

# Interpreting the results

The result table is easiest to read with the simulation truth next to it
(available here only because the data are synthetic):

```{r interpret}
data(postexport_example_truth)
tab <- as.data.frame(res)
cmp <- merge(postexport_example_truth[, c("event", "truth", "sigma_c")],
             tab[, c("event", "sigma_c", "IR", "p_value", "at_boundary")],
             by = "event", suffixes = c("_true", "_est"))
cmp
```

- **`sigma_c`** is the full-model estimate of the post-export conversion
  rate, per minute. It is an effect-size estimate, not a test.
  - Estimates can be positive for events simulated under the null.
  - Estimates can differ strongly from the simulated value. In `alt_3`, for
    example, the full fit trades `sigma_c` against the cytoplasmic decay
    rate `alpha`, since both remove `C`. The manuscript's practical
    identifiability analysis examines this.
- **`IR`**, `(RSS0 - RSS1) / RSS0`, is the relative improvement in fit
  from adding `sigma_c`.
- **`p_value`** is the per-event bootstrap p-value, before any
  multiple-testing adjustment.
- **Boundary status.** When the full fit does not improve on the null
  (`T_obs` at or below the frozen tolerance
  `1e-10 * max(1, |RSS0|, |RSS1|)`), the event is *at the boundary*.
  - `sigma_c` is then typically estimated as 0.
  - The p-value is set to 1 by the frozen boundary rule.
  - Such an event gives no kinetic support for an additional conversion
    component.
- **`status`** reports numerical failures, such as a failed observed fit or
  an unstable bootstrap. Only events with status `"ok"` are valid tests.

In this design only one of the four simulated alternatives has a small
p-value. That is expected: in the manuscript benchmark, this design had
empirical Type-I error close to 0.05 but power of only about 0.24 (see the
design diagnostics below).

# Multiple testing

Adjustment is an explicit step. It is never performed inside the test.
`adjust_postexport_pvalues()`:

- uses Benjamini-Hochberg by default, via `stats::p.adjust()`;
- adjusts only valid tests (status `"ok"` and a finite p-value);
- adds `q_value` and leaves raw p-values unchanged.

```{r adjust}
adj <- adjust_postexport_pvalues(res)
adj$summary[, c("event", "status", "p_value", "q_value")]
```

The testing family should follow the analysis plan. The manuscript adjusted
each dataset separately; `groups` gives one family per label. A single test
can also be adjusted, as a family of size one.

# Exploratory ranking

`rank_postexport_candidates()` orders the tested events by the manuscript's
exploratory composite score

```
score = sigma_c * IR * min(-log10(max(q, 1e-10)), 6)
```

- It requires q-values; raw p-values are never substituted.
- Ranks are computed within each adjustment family.
- Ties share the minimum rank.
- Failed events are kept, with no score.

The ranking is a prioritisation aid for follow-up. It is not an inferential
quantity and not an optimised discrimination score.

```{r rank}
rank_postexport_candidates(adj)
```

# Plots

All plots are `ggplot2` objects. They never refit or modify results.

The fit plot shows the observations, time-point means, sampling times,
`t_star`, and the fitted full and null models. The fitted curves are display
reconstructions, propagated with the ODE model from the observed
replicate-mean state at the first sampled time. Inference does not use them;
it uses the interval-balance / Crank-Nicolson machinery.

```{r plot-fit, fig.height = 6}
plot(res$results$alt_3)
```

The bootstrap plot shows the bootstrap distribution of `T`, the observed
statistic and the fraction of bootstrap replicates at the boundary:

```{r plot-boot}
plot(res$results$alt_3, type = "bootstrap")
```

Set-level plots are descriptive:

- `"status"`, `"boundary"` and `"sigma_IR"`;
- after adjustment, `"sigma_q"`, `"IR_q"` and `"score"`.

```{r plot-set}
plot(adj, type = "sigma_q")
```

# Simulation

`simulate_postexport_kinetics()` exposes the manuscript simulator. It
supports:

- a transcriptional onset and a common intervention time;
- `"NONE"`, `"SHUTOFF"` and `"PSEUDO_SHUTOFF"` regimes;
- replicate-level rate variability;
- the benchmark assay-noise presets (Gaussian, RNA-seq, RT-qPCR).

```{r simulate}
p <- c(R = 100, tau = 0.03, tau_s = 0.015, sigma_c = 0.1, sigma_n = 0.1,
       alpha = 0.2, alpha_s = 0.08)
sim <- simulate_postexport_kinetics(
    p, times = c(90, 100, 110, 120, 130), onset_time = 0, t_star = 100,
    regime = "SHUTOFF", time_unit = "min", n_replicates = 3,
    param_cv = 0.05, noise = list(platform = "rnaseq", level = "low"),
    seed = 1)
sim
plot(sim)
```

`sim$observed` has the wide input format and can be passed directly to
`postexport_data()`. Simulating with `sigma_c > 0` generates data from the
kinetic model; it does not represent any molecular mechanism.

# Operational-domain diagnostics

Calibration of the bootstrap test depends on the design: the regime,
sampling times, replicates and noise. `check_operational_domain()` compares a
design with the designs of the manuscript's corrected factorial benchmark
and reports their empirical Type-I error and power.

```{r domain}
dom <- check_operational_domain(x, regime = "SHUTOFF", t_star = 332,
                                platform = "rnaseq", noise_level = "very_low")
dom
plot(dom)
```

The check is diagnostic only. Similarity to a benchmark design is not a
calibration guarantee for any real dataset, and the check never blocks
inference.

Continuous transcription was strongly anti-conservative in the benchmark:

```{r domain-none}
check_operational_domain(regime = "NONE", n_time_points = 5, n_replicates = 5,
                         sampling_interval = 10, time_unit = "min")
```

# Limitations

- **Phenomenological rate.** `sigma_c` is phenomenological. Statistical
  support for `sigma_c > 0` indicates kinetic patterns consistent with
  additional post-export conversion within the four-state model. It does
  not establish cytoplasmic splicing or any unique molecular mechanism.
- **Design-dependent calibration.** In the manuscript benchmark, Type-I
  error and power varied strongly with the regime, the sampling interval,
  the number of time points and replicates, and the noise. Several designs
  were anti-conservative.
- **Pseudo-shutoff assumptions matter.** The analysis assumes the
  transcription schedule given by `t_star` (complete shutoff here).
  Residual transcription after the intervention, or misassigned fractions,
  can bias the test. Pseudo-shutoff designs are reported as
  `not_benchmarked` by the operational-domain check.
- **Benchmark similarity is not a calibration guarantee.** The benchmark
  used simulated data from the model itself.
- **Continuous transcription.** Without an intervention, the test was
  anti-conservative in the manuscript benchmark.
- **Exploratory ranking.** The composite score is a prioritisation aid, not
  a test.

# Reproducibility and provenance

The numerical core reproduces the frozen manuscript implementation:

- tag `manuscript-revision-v1.0`, commit
  `65c3b7368fb7686bfde3dab857f98c393bb534c5`, in
  <https://github.com/bioinformatics-sannio/postexport-kinetics>;
- archived at <https://doi.org/10.5281/zenodo.22944109>. This DOI identifies
  the frozen manuscript implementation, not this package, which has no DOI.

Results are bitwise reproducible only within a matched numerical
environment. The bootstrap draws use `MASS::mvrnorm()`, whose output depends
on the LAPACK/BLAS build. Across platforms, bootstrap p-values are
different Monte-Carlo realisations of the same bootstrap distribution. The
deterministic fit quantities agree to rounding level.

Use `citation("postexportKinetics")` to cite the package and the
manuscript.

# Session information

```{r session}
sessionInfo()
```
